---
title: "Homework 3"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Question 1) 

```{r Question 1) Setting up the given data}

if (!require('tidyverse')) 
{
  library("tidyverse")
  install.packages("dplyr")
  library("dplyr")
}

```

```{r Question 1 Given Data )}

# Pulling the tables and setting up the data
table2
table4a
table4b

# Fidning the length of the table
length = nrow(table2)

```

```{r Part A Table 2}

# Filtering table 2 to create a table with only the cases
table2_cases <- filter(table2,type == 'cases') 
View(table2_cases)

```

```{r Part B Table 2}

# Filtering table 2 to create a table with only the population
table2_population <- filter(table2,type == 'population')
View(table2_population)

```

```{r Part C and D Table 2}

# Joining the tables together by country and year
table2_cases_population <- left_join(table2_cases,table2_population, by=c("country","year"))

# Calculating the rate, while simultaneously storing the data points within a new column
table2_cases_population_rate <- table2_cases_population %>% mutate(rate = (count.x/count.y) *10000)

# Renaming the columns
table2_cases_population_rate <- table2_cases_population_rate %>% rename(cases = count.x)
table2_cases_population_rate <- table2_cases_population_rate %>% rename(population = count.y)

# Viewing the table
table2_cases_population_rate

```

```{r Part A and B for Table 4A and Table4B}

# Gathering the given data by cases for the years 1999-2000
table4a_cases <- gather(table4a,"year","cases", "1999":"2000")
View(table4a_cases)

# Gathering the given data by population for the years 1999-2000
table4b_population <- gather(table4b,"year","population", "1999":"2000")
View(table4b_population)

```

```{r Part C and D for Tables 4A and 4B}

# Merging the tables along country and year
table4a_4b_merge <- left_join(table4a_cases,table4b_population, by=c("country","year"))

# Calculating the rate, while simultaneously storing the data points within a new column
table4a_4b_merge_rate <- table4a_4b_merge %>% mutate(rate = (cases/population) *10000)

# Viewing the table
View(table4a_4b_merge_rate)

```

Therefore, there are two ways to calculate the rate of the disease for the 3 countries per year. The first method indicated how the table 2 data can be filtered and subsequently mutated to find the rate of the disease (tuberculosis) per year for the different countries. On the other hand, the second method combines the data from two tables (4a and 4b) which cantain information regarding either the population or the cases for a country by year. Therefore, the data has to be combined and gathered from the two tables and then mutated. Both these methods have been illustrated above.

# Question 2) 

```{r Question 2}

#table4a %>% gather(1999,2000,key="year",value="cases")

```
The error message produced by this code block is as follows: "Error: Can't subset columns that don't exist. x Locations 1999 and 2000 don't exist. There are only 3 columns"; Therefore, it becomes apparent from looking at documentation concerning the gather function, that 4 arguments were provided concerning only 3 columns. As a result, this error was produced. Here the gather function within the dplyr library needs to be used in the following way, with the appropriate syntax: 

```{r Question 2 Corrected Syntax}

table4a %>% gather(key="year",value="cases", "1999":"2000")

```
Looking at the syntax above, it becomes clear why the error message was initially displayed. With the original syntax, errors occured as the code was attempting to subset into columns '1999' and '2000'. As these columns did not exist an error was returned. On the other hand, with the corrected syntax, the data points were gathered for 'year' and 'cases' between the values 1999-2000. Therefore, the original code fails, while the corrected code succeeds. 

# Question 3) 

```{r Question 3) Setting up the given data}

if (!require('nycflights13') | !require('dplyr') | !require('plyr') ) 
{
  library("nycflights13")
  install.packages("dplyr")
  library("dplyr")
  install.packages("plyr")
  library("plyr")
  install.packages("ggplot2")
  library("ggplot2")
}

```

```{r Question 3 Part A}

# Gathering the given data
flights

# Grouping the flights data within a new data set
flights_df <- flights %>%
  group_by(year, month, day)

# Selecting columns to work with
flights_df <- flights_df %>%
  select(year, month, day, dep_time, sched_dep_time, dep_delay, air_time, hour, minute, time_hour)

# Uniting and converting into date columns
flights_df <- flights_df %>%
  unite("dep_day_edited", year, month, day) %>%
  mutate(dep_day_edited = as.Date(dep_day_edited, format="%Y_%m_%d"))

# Converting column type into integer
flights_df <- transform(flights_df, air_time=as.integer(air_time))

# Finding the mean flight time per hour
flights_3a <- ddply(flights_df, .(time_hour), summarize, mean = mean(air_time))

# Extracting only the unique values
flights_3a <- unique(flights_3a)

# Plotting the trends observed
plot(flights_3a$time_hour, flights_3a$mean)


```

```{r Question 3 Part B}
# To prove that dep_time, sched_dep_time and dep_delay are consistent I need to verify that the dep_time is always equal to the sum of the sched_dep_time and dep_delay

# Created a new variable called ver_dep_time 
flights_v <- flights %>% 
  mutate(ver_dep_time = sched_dep_time + dep_delay)

# Filter for those rows where dep_time and ver_dep_time match
flights_vr <- flights_v %>% 
  filter(ver_dep_time == dep_time)

# Count of mismatches
count_mismatch = nrow(flights_v) - nrow(flights_vr)
count_mismatch

# Filter for those rows where dep_time and ver_dep_time do not match 
flights_ur <- flights_v %>% 
  filter(ver_dep_time != dep_time)

flights_ur
# Therefore, it becomes apparent that there are discrepancies between the dep_time, sched_dep_time and dep_delay for a large section of the values. Here the dep_time does not match the sum of the dep_delay with the sched_dep_time; As a result, these columns are not fully consistent with one another. 
```

```{r Question 3 Part C Hypothesis Confirmation}
# To start, I created a binary variable that will be used to help tell me if a flight was delayed or not
 flights_b <- flights %>% 
  mutate(delayed = case_when(dep_delay < 0 ~ 1, 
                             dep_delay > 0 ~ 0))
  
# Next I grouped the data by minutes
flights_b <- flights_b %>% 
  group_by(minute)
  
# I removed NA values
flights_b <- flights_b %>% 
  filter(!is.na(delayed))

# Filtered the data for minutes between 20:30 and 50:60
flights_b <- filter(flights_b, (minute == (20:30)) | (minute == (50:60)))

# I found the percent of early departure for each minute of departure. Here, the flights_sch table indicates the proportion of flights that left early for each minute of departurre bwteen 20 and 30 and 50 and 60  

flights_s <- ddply(flights_b, .(time_hour), summarize, mean = mean(delayed))

flights_s <- transform(flights_s, round(mean=as.numeric(mean)),2)

# Sample Alternative Code
# flights_s <- flights_b %>% summarise(percentage = mean(delayed))

# Visual Representation of the data
#plot(flights_s$minute, flights_s$percentage)

# Finding the percentage of delay for flights 

count_1_df <-  flights_b %>% 
  filter(delayed == 1)

count_1 <- nrow(count_1_df)

count_0_df <-  flights_b %>% 
  filter(delayed == 0)

count_0 <- nrow(count_0_df)
count_0 + count_1

percentage = (count_1/nrow(flights_b))*100
percentage

# Confirming that those early departures between minutes 20-30 and 50-60 are caused by scheduled flights that leave early
flights_sch <- filter(flights_b, (dep_delay < 0))
flights_sch

# Therefore, it becomes apparent that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Therefore, the hypothesis is confirmed. 

```


