---
title: "Data Wrangling Final"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Rahul Babu
# QBS 181 Final

# Question 1 Blood Pressure and Demographics Data 

```{r Setting up a working directory}

# Removing information regarding personal directoy before publishing
setwd("")

```

```{r Question 1 Loading File }

data <- read.csv(file = 'IC_BP_v2.csv')

```

# Part A) Convert BP alerts to BP status

```{r Question 1 Part A Loading Packages}

if (!require('dplyr')) 
{
  install.packages("dplyr")
  library("dplyr")
  install.packages("lazyeval")
  library("lazyeval")
}

if (!require('tidyr')) 
{
  install.packages("tidyr")
  library("tidyr")
}

```

```{r Question 1 Part A}

# Creating 2 Data Sets
data_2 <- data

# Creating and copying a new column for BP Status 
#data$BPStatus <- data$BPAlerts

# Checking the new data set 
#head(data, 10)

# Converting BPAlerts Column name to BPStatus 
data_2 <- data_2 %>% rename(BPStatus = BPAlerts)

# Printing Top 10 rows to show this change
head(data_2, 10)

```

```{r Question 1 Part B}

# Converting into a data frame
as.data.frame(data_2)

# Using case when to create a new Outcomes column
data_2 <- data_2 %>% 
  mutate(Outcomes = case_when(data_2$BPStatus == "Hypo1" | data_2$BPStatus == "Normal" ~ "Controlled", 
                             data_2$BPStatus == "Hypo2" | data_2$BPStatus == "HTN1" | data_2$BPStatus == "HTN2" | data_2$BPStatus == "HTN3" ~ "Uncontrolled", 
                             data_2$BPStatus == "NULL" ~ "Unknown"))

# Creating a Dichotomous Outcome Table
data_2 <- data_2 %>% mutate(Dich = case_when(data_2$Outcomes == "Controlled" ~ 1,
                                             data_2$Outcomes == "Uncontrolled" ~ 0
                                             ))

# Removing NA rows with respect to the Dichotomous Outcome Table (i.e. Unknown Outcome rows were removed)
data_2 <- data_2[!is.na(data_2$Dich), ]

# Printing top 10 rows to show these changes
head(data_2, 10)

# Checking no. of rows that were removed
removed_number = nrow(data) - nrow(data_2)

```

```{r Question 1 Part C Loading Packages}

if (!require('RODBC')){
  install.packages('sqldf')
  install.packages('RODBC')
  library('sqldf')
  library('RODBC')
  install.packages('ddply')
  library('plyr')
}

```

```{r Question 1 Part C}

# Establishing my connection
# Removing my password and username before publishing
my_connection <- odbcConnect('qbs181', uid = '', pwd = '',)

# Pulling the Demographics Data 
demographics_data <- sqlQuery(my_connection, "SELECT * FROM Demographics")

# Merging the tables
merge_table <- merge(data_2, demographics_data, by.x = "ID", by.y = "contactid")

# Printing out top 10 rows to show these changes
head(merge_table, 10)

```

```{r Question 1 Part D}

# Converting into a data frame
merge_table <- as.data.frame(merge_table)

# Converting columns in dates
merge_table$tri_imaginecareenrollmentemailsentdate <- as.Date(merge_table$tri_imaginecareenrollmentemailsentdate, format = "%m/%d/%Y")
merge_table$tri_enrollmentcompletedate <- as.Date(merge_table$tri_enrollmentcompletedate, format = "%m/%d/%Y")

# Computing the time of enrollment per row
merge_table$tri_enrollmentlength <- as.Date(merge_table$tri_enrollmentcompletedate, format = "%m/%d/%Y") - as.Date(merge_table$tri_imaginecareenrollmentemailsentdate, format = "%m/%d/%Y")

# How many rows had enrollment above 12 weeks? 
length(which(merge_table$tri_enrollmentlength > 83))

# Computing the mean dichotomous per week (per observed group) per patient
means_dichotomous <- ddply(merge_table,c("ID","ObservedTime"),summarise, mean_dich_week =mean(Dich))
merge_table <- merge(merge_table, means_dichotomous, by = c("ID","ObservedTime"))

# Computing the mean dichotomous total per patient
means_dichotomous_total <- ddply(merge_table,c("ID"),summarise, mean_dich_total =mean(Dich))
merge_table <- merge(merge_table, means_dichotomous_total, by = c("ID"))

# Computing the mean systolic value per patient
means_systolic_total <- ddply(merge_table,c("ID"),summarise, avg_systolic =mean(SystolicValue))
merge_table <- merge(merge_table, means_systolic_total, by = c("ID"))

# Computing the mean diastolic value per patient
means_diastolic_total <- ddply(merge_table,c("ID"),summarise, avg_diastolic =mean(Diastolicvalue))
merge_table <- merge(merge_table, means_diastolic_total, by = c("ID"))

# Ordering by contact id and observed time
merge_table <- merge_table[with(merge_table, order(ID,ObservedTime)), ]

# Printing out the top 10 rows to show the changes
head(merge_table, 10)

```

```{r Question 1 Part E}

# Filtering data for those patients observed for more than 83 days ~ for 12 weeks
merge_table_12_weeks <- filter(merge_table, tri_enrollmentlength > 83)

# Finding number of patients enrolled for more than 12 weeks 
patients_12_week_count <- length(unique(merge_table_12_weeks$ID))

# Obtaining only the first row per ID (finding the first recorded result ordered by observed time) for those patients >= 12 weeks
merge_table_top <- merge_table_12_weeks %>% group_by(ID) %>% slice(1)

# Obtaining only the last row per ID (finding the last recorded result ordered by observed time) for those patients >= 12 weeks
merge_table_bot <- merge_table_12_weeks %>% group_by(ID) %>% slice(n())

# Creating a comparative table 
merge_table_bot$Dich_baseline <- merge_table_top$Dich
  
# Creating a comparative table 
comparative <- merge_table_bot

# Printing out top 10 rows to show the changes 
head(comparative, 10)

# Obtaining only the first row per ID (finding the first recorded result ordered by observed time) for all patients
merge_table_top_total <- merge_table %>% group_by(ID) %>% slice(1)

# Obtaining only the last row per ID (finding the last recorded result ordered by observed time) for all patients
merge_table_bot_total <- merge_table %>% group_by(ID) %>% slice(n())

# Creating a comparative table 
merge_table_bot_total$Dich_baseline <- merge_table_top_total$Dich
  
# Creating a comparative table 
comparative_total <- merge_table_bot_total

# Printing out the top 10 rows to show the changes 
head(comparative_total, 10)

```
There are two tables displayed above, comparative and comparative total. Comparative compares for only those patients enrolled for more than 83 days (for atleast 12 weeks). On the other hand, comparative total compares for all patients within the data set. 

```{r Question 1 Part F}

# Finding the number of Controlled and Uncontrolled Patients at week 1
uncontrolled_week_1 = length(which(merge_table_top$Outcomes=="Uncontrolled"))
controlled_week_1 = length(which(merge_table_top$Outcomes=="Controlled"))

# Finding the number of Controlled and Uncontrolled Patients at end of enrollment week 12
uncontrolled_week_12 = length(which(merge_table_bot$Outcomes=="Uncontrolled"))
controlled_week_12 = length(which(merge_table_bot$Outcomes=="Controlled"))

# Finding the number of patients who went from Uncontrolled to Controlled for those enrolled for 12 weeks
uncontrolled_week_1 - uncontrolled_week_12

# Finding the number of Controlled and Uncontrolled Patients at week 1
uncontrolled_total_1 = length(which(merge_table_top_total$Outcomes=="Uncontrolled"))
controlled_total_1 = length(which(merge_table_top_total$Outcomes=="Controlled"))

# Finding the number of Controlled and Uncontrolled Patients at end of enrollment week 12
uncontrolled_total_12 = length(which(merge_table_bot_total$Outcomes=="Uncontrolled"))
controlled_total_12 = length(which(merge_table_bot_total$Outcomes=="Controlled"))

# Finding the number of patients who went from Uncontrolled to Controlled for all those enrolled
uncontrolled_total_1 - uncontrolled_total_12

```     

Therefore, for those patients observed for over 12 weeks, 1 patient moved from uncontrolled to controlled. On the other hand, when considering all the patients, regardless of their period of enrollment, 7 patients moved from uncontrolled to controlled. 

# Question 2: Merging Tables in SQL 

This question and its resulting output was done in SQL and will be uploaded separately

# Question 3: Merging Tables in R

```{r Question 3 Loading Packages}

if (!require('data.table')){
  library(data.table)
}

```

```{r Question 3}

# Pulling tables from sql server using ODBC connection
demographics <- sqlQuery(my_connection, "SELECT * FROM Demographics")
conditions <- sqlQuery(my_connection, "SELECT * FROM Conditions")
messages <- sqlQuery(my_connection, "SELECT * FROM Text")

# Merging the tables along contactid
R_merge <- merge(demographics, conditions, by.x = "contactid", by.y = "tri_patientid")
R_merge <- merge(R_merge, messages, by.x = "contactid", by.y = "tri_contactId")

# Converting to Date time
R_merge$TextSentDate <- as.Date(R_merge$TextSentDate, format = "%Y-%m-%d")
R_merge$tri_imaginecareenrollmentemailsentdate <- as.Date(R_merge$tri_imaginecareenrollmentemailsentdate, format = "%m/%d/%Y")
R_merge$tri_enrollmentcompletedate <- as.Date(R_merge$tri_enrollmentcompletedate, format = "%m/%d/%Y")

# Using setDT() as I am working with large amounts of data in an effort to conserve memory
R_merge_edited <- setDT(R_merge)

# Finding the max text sent date time per contactid
# Here, .SD is used to subset or self-reference the data
R_merge_edited <- R_merge_edited[, .SD[which.max(TextSentDate)], by = contactid]

# Checking 
nrow(R_merge_edited) == length(unique(R_merge$contactid))

# Printing top 10 and bottom 10 rows to show this change
head(R_merge_edited, 10)

```

# Question 4

My repository can be found under the following name: https://github.com/rbabu98/Data_Wrangling_Projects_and_Tasks

